\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields

### 2023-10-18 doc/paper-jss/sdmTMB-paper.Rnw
Restructure user interface section

@@

Here, we introduce the \R\ package \pkg{sdmTMB}, which implements geostatistical spatial and spatiotemporal GLMMs using \pkg{TMB} for model fitting and \INLA\ to set up SPDE matrices.
Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} \citep{wood2017a} for a specific class of spatial and spatiotemporal models.
Many features of \pkg{sdmTMB} may be found in other software (Table~\ref{tab:functionality}), but to date no package has included all in a single package.
One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \pkg{TMB}: \pkg{sdmTMB}), although the package is widely applicable to other fields and any geostatistical data collected continuously in space and discretely in time.

@@

\section{Software}\label{sec:software}

The design goal of \sdmTMB\ was to develop a flexible, modular, and intuitive inteface to fast maximum likelihood inference (or full Bayesian inference) with the SPDE approach to spatial and spatiotemporal GLMMs with random fields.
The package gathers functionality not found combined in other packages that is particularly useful to species distribution modelling, but is applicable beyond ecology to any field that encounters geostatistical data that is continuously referenced in space and (optionally) discretely indexed by time.

\sdmTMB\ relies on several well-established \R\ packages (Figure~\ref{fig:flowchart}).
\sdmTMB\ uses \pkg{fmesher} \citep{fmesher} and \INLA\ \citep{lindgren2015} to generate the triangulation mesh and input matrices for the SPDE calculations.
The SPDE precision matrix calculations themselves are performed by functions within \pkg{TMB} \citep{kristensen2016} (\texttt{R\_inla} namespace).
Data structures required to implement penalized smoothers are constructed with \texttt{smooth2random()} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
The statistical model itself is written as a \proglang{C++} template for \TMB.
\TMB\ returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
The negative marginal log likelihood is minimized via the non-linear optimization routine \texttt{stats::nlminb()} \citep{gay1990} in \proglang{R} with additional optimization carried out via a Newton optimization step to find the Hessian using \texttt{stats::optimHess()} \citep{gay1990, r2021}.
After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to samples from the joint posterior distribution for Bayesian inference.
% TODO: visreg + ggeffects effects, etc.?

@@

The \sdmTMB\ package is designed to be both modular and familiar to users of widely used \R\ packages (e.g., \pkg{glmmTMB}, \pkg{lme4}, \pkg{mgcv}).
The user starts by constructing a triangulation mesh for the SPDE approach with \texttt{make\_mesh()}.
\texttt{make\_mesh()} is a wrapper for \pkg{fmesher} and \INLA's triangulation mesh functions and users can also construct any mesh with \pkg{fmesher} or \INLA\ and pass it to \texttt{make\_mesh()}.
Fitting is accomplished with \texttt{sdmTMB()}, which has arguments similar to \pkg{glmmTMB}'s \texttt{glmmTMB()} but with additional arguments for how any spatial and spatiotemporal random fields are structured, what column defines time, any time-varying formulas, and any spatially varying formulas.
Families are specified with standard R family functions (e.g., \texttt{binomial()}) plus several \sdmTMB-specific families for family's not available in the \R\ stats library (e.g., \texttt{nbinom2()}, \texttt{delta\_lognormal()}, see \texttt{?sdmTMB::Families}).
The formula syntax follows a standard \R\ format and is expanded on in the next paragraph.
A fitted model summary can be viewed with \texttt{print.sdmTMB()} or \texttt{summary.sdmTMB()} and a set of basic ``sanity'' checks can be run with \texttt{sanity()}.
\texttt{tidy()} returns parameter estimates in standard data frame formats similar to \pkg{broom} \citep{broom}.
Others standard S3 methods are also available such as \texttt{fixef()}, \texttt{confint()}, and \texttt{vcov()}.
Prediction on fitted or new data is accomplished with \texttt{predict.sdmTMB()}.
\texttt{residuals.sdmTMB()} calculates various types of residuals.
New observations can be simulated from a fitted model with \texttt{simulate.sdmTMB()} or new data can be simulated without a fitted model (``de novo'') with \texttt{sdmTMB\_simulate()}.
\texttt{sdmTMB\_cv()} facilitates cross validation.
Derived quantities can be calculated on the fitted model.
% TODO: more COG refs
Common quantities of interest include summarizing shifts in the center of gravity (\texttt{get\_cog()}) \citep[e.g.,][]{thorson2016cog}, or tracking total densities through time (\texttt{get\_index()}).
% The \texttt{get\_index()} function generates a weighted average of densities for the prediction grid, integrating across space; for prediction grids with equally sized cells, each cell will contribute equally to the total average, however cells may be of unequal size.

@@

The \sdmTMB\ formula syntax (\texttt{formula}) follows a standard format that is similar to \pkg{glmmTMB}, \pkg{lme4}, and \pkg{mgcv}.
In addition to standard main effects, the user can include random intercepts (e.g., \texttt{+ (1 | group)}), threshold shaped hockey-stick models \texttt{breakpt()} \citep{barrowman2000}, logistic functions \texttt{logistic()}, and smoothers.
Penalized smoothers (i.e., for generalized additive models, GAMs) use the same \texttt{s()} and \texttt{t2()} syntax as in \pkg{mgcv} \citep{wood2017a}.
Supported functionality includes bivarate smoothers \texttt{s(x, y)},
smoothers varying by continuous or categorical variables \texttt{s(x1, by = x2)},
cyclical smoothers \texttt{s(x, bs = "cc")},
and smoothers with specified basis dimensions \texttt{s(x, k = 4)} \citep{wood2017a}.
In addition to the main formula, \texttt{sdmTMB()} accepts \emph{one-sided} formulas for coefficients that should vary through time (\texttt{time\_varying}) according to a random walk or AR(1) process (\texttt{time\_varying\_type}) or vary through space as random fields (\texttt{spatial\_varying}).

The \texttt{predict.sdmTMB()} method is flexible and includes the option to specify a new data frame (\texttt{newdata}),
whether to return predictions on the link or response scale (type),
whether to return standard errors (\texttt{se\_fit},
can be slow if conditioned on random fields),
whether to condition on the random fields (\texttt{re\_form}),
whether to condition on the random intercepts (\texttt{re\_form\_iid}),
which delta model linear predictor to use (\texttt{model}),
whether to take draws from the joint parameter precision matrix (\texttt{nsim}),
and whether to use MCMC samples from a \pkg{tmbstan} model fit (\texttt{mcmc\_samples}).

\sdmTMB\ models can include penalized likelihoods by assigning priors (penalties) to model parameters through the \texttt{sdmTMB()} \texttt{prior} argument (\texttt{?sdmTMBpriors}).
These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
Following other recent SPDE implementations in \pkg{TMB} \citep{breivik2021, osgood-zimmerman2021}, penalized complexity (PC) priors \citep{simpson2017, fuglstad2019} (\texttt{?pc\_matern}) can constrain the spatial range and variance parameters.
These priors or penalties are available both with maximum likelihood estimation and with MCMC sampling.
If a model will be passed to \pkg{tmbstan}, a \texttt{bayesian} logical argument is available to enable Jacobian adjustments for changes of variables (priors applied to parameters that are internally transformed) \citep{carpenter2017}.

@@

\subsection{Installation}\label{installation}

@@

<<sdmTMB-install, eval=FALSE, echo=TRUE>>=
install.packages("sdmTMB", dependencies = TRUE)
@

Importantly, \texttt{dependencies = TRUE} includes the \INLA\ package \citep{rue2009, lindgren2011, lindgren2015}, which is not on CRAN but is required to generate matrices for the SPDE.
%Soon, we intend to switch to using the \pkg{fmesher} \citep{fmesher} package to eliminate this dependency.

Alternatively, the development version can be installed with:

<<remotes, eval=FALSE, echo=TRUE>>=
# install.packages("remotes")
remotes::install_github("pbs-assess/sdmTMB", dependencies = TRUE)
@

@@

We can test for spatial autocorrelation with a Moran's I test or with a visual inspection of the residuals calculated from the \texttt{residuals()} S3 method.
We omit example code for Moran's I calculation for brevity but include it in the code supplement.
The p-value for spatial autocorrelation is low for the model without a random field (p $<$ 0.01) but large for the model with a random field (p = 0.84) suggesting significant residual correlation in the simpler model that is alleviated by including the random field.

@@

\sloppy To visualize our model, we can make predictions with the \texttt{predict()} method (\texttt{?predict.sdmTMB}) and optionally use the \texttt{newdata} argument to predict on a new data frame with any locations and values for the predictor columns.
Here, we will predict on a 2 \(\times\) 2 km grid (\texttt{qcs\_grid}) that covers the entire region of interest so we can visualize the predictions spatially.
The grid contains spatial covariate columns and all predictors used in the model set at values for which we want to predict.
The output of \texttt{predict()} is a data frame containing overall estimates in link space (\texttt{est}), estimates from the non-random-field components (\texttt{est\_non\_rf}; here, intercept and depth), and estimates from the individual random field components (here, \texttt{omega\_s}---the spatial field).
We can plot these with \texttt{geom\_raster()} from the \pkg{ggplot2} package (Figure~\ref{fig:pcod-predict-maps}).

@@

<<dog-wcvi-pred, fig.asp = 0.8, fig.cap="Example prediction components from the spatiotemporal model of Pacific Dogfish biomass density. Throughout, two example years are shown. \\code{est\\_non\\_rf1} refers to the prediction from all non-random-field components (here, a smoother for bottom depth and the time-varying year effect) from the first delta model, \\code{omega\\_s1} refers to the spatial random field from the first delta model, \\code{epsilon\\_st2} refers to spatiotemporal random fields from the second delta model, and \\code{est} refers to the overall prediction estimate combining all effects. The spatial random field is constant through time and represents static biotic or abiotic features not included as covariates (e.g. habitat). The spatiotemporal random fields are different each time step and here are allowed to follow an AR(1) process. They represent temporal variability in the spatial patterning of Pacific Spiny Dogfish (e.g., resulting from movement or local changes in population density).", fig.width=9, out.width="6in">>=
cowplot::plot_grid(
  g_nonrf,
  g_omega,
  g_eps,
  g_est,
  ncol = 2L
)
@

@@

There are many \proglang{R} packages capable of fitting geostatistical spatial or spatiotemporal models \citep[e.g.,][]{heaton2019}.
\pkg{sdmTMB}, \pkg{VAST}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, and \pkg{spaMM} \citep{rousset2014} are the most closely related, as they all provide a user interface to SPDE-based GRF models.
In our software comparison (Table \ref{tab:functionality}), we also include \pkg{mgcv} as it can be adapted to use the SPDE \citep{miller2019} and \pkg{spBayes} \citep{finley2007, finley2015} since it is a prominent package that can fit related predictive-process models without the SPDE.
\pkg{sdmTMB}, \pkg{VAST}, and \pkg{mgcv} can estimate anisotropic covariance whereas \proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{spBayes} are currently limited to isotropic covariance.
\pkg{sdmTMB} and \pkg{mgcv} focus on univariate response data, whereas \pkg{VAST}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, \pkg{spaMM}, and \pkg{spBayes} extend to multivariate responses with various limitations.
To our knowledge, \pkg{VAST} is the only package to implement spatial \citep{thorson2015} and spatial dynamic factor analysis \citep{thorson2016} and spatial empirical orthogonal function (EOF) regression \citep{thorson2020}.
Of these packages, only \pkg{sdmTMB} and \pkg{inlabru} can currently fit threshold (e.g., hockey-stick) covariate relationships.
\pkg{spaMM} is limited to a spatial random field (i.e., does not fit spatiotemporal fields) and \pkg{spBayes} implements spatiotemporal fields, but only as a random walk.
There is considerable variability in the available observation likelihoods across packages (Table \ref{tab:functionality}).
% We provide comparisons of the syntax and the reproducibility of results from models fit using \pkg{sdmTMB}, \proglang{R}-\pkg{INLA}, and \pkg{inlabru} in Appendix~\ref{app:inla}.

@@

% TODO: figure out where to put this paragraph
The GLMMs underpinning \pkg{sdmTMB} models are spatially explicit---they estimate interpretable parameters of a spatial covariance function: parameters defining the magnitude of spatial variation and the rate of correlation decay with distance.
In contrast, semi- and non-parametric approaches do not estimate spatial covariance functions (e.g., \pkg{randomForest} \citep{liaw2002}, \pkg{MaxEnt} \citep{phillips2006}, and most smooths in \pkg{mgcv} \citep{wood2017a}).
The random fields in \pkg{sdmTMB} are structured as MVN constrained by a Matérn covariance function \citep{matern1986}.
The Matérn can accommodate a range of shapes and can be both isotropic (correlation decays the same in all directions) or anisotropic (correlation is directionally dependent) \citep{haskard2007}.
The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to \(\sim 0.13\) \citep{lindgren2015}---can be shared among random fields or estimated separately (\texttt{share\_range} argument).

% TODO: figure out where to put this paragraph
Areal data (data aggregated to polygon or grid level) may be analyzed using other spatial models, including conditional (CAR) autoregressive models \citep[e.g.,][]{verhoef2018}; \pkg{sdmTMB} can also fit models with areal data if each polygon has an associated centroid.
A benefit of the geostatistical approach over CAR or similar models is that the parameters describing spatial covariance can be more easily interpreted \citep{wall2004} (e.g., spatial variance and range---the distance at which points are effectively independent).

@@

There are several planned future additions to the \pkg{sdmTMB} model structure.
A subset of features to be added in the near future includes (1) multivariate models to allow for simultaneous modelling of data from multiple species, (2) integration with the \pkg{RTMB} \citep{rtmb} package so that the model code base will be written in \proglang{R} rather than \proglang{C++}, (3) inclusion of a dispersion formula similar to \pkg{glmmTMB}, (4) continuous-time models \citep[e.g.,][]{blangiardo2015book}, (5) non-Gaussian random fields \citep[e.g.,][]{anderson2019}, and (6) multivariate responses.
The included \pkg{TMB} \texttt{.cpp} file is flexible and provides a tested model template that can be modified to add additional features.
% TODO random slopes looks gnarly to do generically? EW: -- agree, I mostly have looked at glmmTMB / brms implementations
% removed random slopes -- another feature that could be added is 3D models, e.g. https://doi.org/10.1016/j.spasta.2023.100750

>>>>

% setwd("doc/paper-jss/")
% knitr::knit("sdmTMB-paper.Rnw");tinytex::latexmk("sdmTMB-paper.tex", engine = "xelatex")

@@

Here, we introduce the \R\ package \sdmTMB, which implements geostatistical spatial and spatiotemporal GLMMs using \pkg{TMB} for model fitting and \INLA\ to set up SPDE matrices.
Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} \citep{wood2017a} for a specific class of spatial and spatiotemporal models.
Many features of \sdmTMB\ may be found in other software (Table~\ref{tab:functionality}), but to date no package has included all in a single package.
One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \pkg{TMB}: \pkg{sdmTMB}), although the package is widely applicable to other fields and any geostatistical data collected continuously in space and discretely in time.

@@

\section{User interface}\label{sec:software}

The design goal of \sdmTMB\ was to develop a flexible, modular, and intuitive interface to fast maximum likelihood inference (or full Bayesian inference) with the SPDE approach to spatial and spatiotemporal GLMMs with random fields.
The package gathers functionality not found combined in other packages that is particularly useful to species distribution modelling, but is applicable beyond ecology to any field that encounters geostatistical data that is continuously referenced in space and (optionally) discretely indexed by time.
\sdmTMB\ relies on several well-established \R\ packages to construct and fit models (Figure~\ref{fig:flowchart}).

@@

The \sdmTMB\ package is designed to be both modular and familiar to users of widely used \R\ packages (e.g., \pkg{glmmTMB}, \pkg{lme4}, \pkg{mgcv}).
The user starts by constructing a triangulation mesh for the SPDE approach with \texttt{make\_mesh()} (Figure~\ref{fig:flowchart}).
\texttt{make\_mesh()} is a wrapper for \pkg{fmesher} \citep{fmesher} triangulation mesh functions and users can also construct any mesh with \pkg{fmesher} or \INLA\ and pass it to \texttt{make\_mesh()}.

Fitting is accomplished with \texttt{sdmTMB()}, which has arguments similar to \pkg{glmmTMB}'s \texttt{glmmTMB()} but with additional arguments for how any spatial and spatiotemporal random fields are structured, what column defines time, any time-varying formulas, and any spatially varying formulas.
Observation distributions and links are specified with standard \R\ family functions (e.g., \texttt{binomial()}) plus several \sdmTMB-specific families not available in the \R\ stats library (e.g., \texttt{nbinom2()}, \texttt{delta\_lognormal()}, see \texttt{?sdmTMB::Families}).

The \sdmTMB\ formula syntax (\texttt{formula} argument) follows a standard format that is similar to \pkg{glmmTMB}, \pkg{lme4}, and \pkg{mgcv}.
In addition to standard main effects, the user can include random intercepts (e.g., \texttt{+ (1 | group)}), threshold shaped hockey-stick models \texttt{breakpt()} \citep{barrowman2000}, logistic functions \texttt{logistic()}, and penalized smoothers \texttt{s()} for generalized additive models, GAMS \citep{wood2017a}.
Penalized smoothers use the same \texttt{s()} and \texttt{t2()} syntax as in \pkg{mgcv} \citep{wood2017a}.
Supported functionality includes bivarate smoothers \texttt{s(x, y)},
smoothers varying by continuous or categorical variables \texttt{s(x1, by = x2)},
cyclical smoothers \texttt{s(x, bs = "cc")},
and smoothers with specified basis dimensions \texttt{s(x, k = 4)} \citep{wood2017a}.
Beyond the main formula, \texttt{sdmTMB()} accepts one-sided formulas for coefficients that should vary through time (\texttt{time\_varying}) according to a random walk or AR(1) process (\texttt{time\_varying\_type}) or vary through space as random fields (\texttt{spatial\_varying}).

Once the user makes a call to \texttt{sdmTMB()}, input data structures for a \TMB\ model are constructed internally (Figure~\ref{fig:flowchart}).
If needed, data structures required to implement penalized smoothers are constructed with \texttt{smooth2random()} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
\texttt{sdmTMB()} formats data, establishes parameter starting values, and constructs an objective function with derivatives based on a compiled \proglang{C++} template written for \TMB.
The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
The negative marginal log likelihood is minimized via the non-linear optimization routine \texttt{stats::nlminb()} \citep{gay1990} in \proglang{R} with additional optimization carried out via a Newton optimization step to find the Hessian using \texttt{stats::optimHess()} \citep{gay1990, r2021}.
After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to samples from the joint posterior distribution for Bayesian inference.

A fitted model summary can be viewed with \texttt{print()} or \texttt{summary()} and a set of basic ``sanity'' checks can be run with \texttt{sanity()}.
\texttt{tidy()} returns parameter estimates in standard data frame formats similar to \pkg{broom} \citep{broom}.
Others standard S3 methods are also available such as \texttt{fixef()}, \texttt{confint()}, and \texttt{vcov()}.

Prediction on fitted or new data is accomplished with \texttt{predict()} (\texttt{?predict.sdmTMB}).
The \texttt{predict.sdmTMB()} method is flexible and includes the option to specify a new data frame (\texttt{newdata}),
whether to return predictions on the link or response scale (type),
whether to return standard errors (\texttt{se\_fit},
can be slow if conditioned on random fields),
whether to condition on the random fields (\texttt{re\_form}),
whether to condition on the random intercepts (\texttt{re\_form\_iid}),
which delta model linear predictor to use (\texttt{model}),
whether to take draws from the joint parameter precision matrix (\texttt{nsim}),
and whether to use MCMC samples from a \pkg{tmbstan} model fit (\texttt{mcmc\_samples}).

A variety of model evaluation tools are available.
A \texttt{residuals()} method calculates various types of residuals.
New observations can be simulated from a fitted model with \texttt{simulate()} or new data can be simulated without a fitted model (``de novo'') with \texttt{sdmTMB\_simulate()}.
\texttt{sdmTMB\_cv()} facilitates cross validation.

% Derived quantities can be calculated on the fitted model.
% TODO: more COG refs
% Common quantities of interest include summarizing shifts in the center of gravity (\texttt{get\_cog()}) \citep[e.g.,][]{thorson2016cog}, or tracking total densities through time (\texttt{get\_index()}).

% The \texttt{get\_index()} function generates a weighted average of densities for the prediction grid, integrating across space; for prediction grids with equally sized cells, each cell will contribute equally to the total average, however cells may be of unequal size.

@@

\pkg{sdmTMB} models can include penalized likelihoods by assigning priors (penalties) to model parameters through the \texttt{sdmTMB()} \texttt{prior} argument (\texttt{?sdmTMBpriors}).
These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
Following other recent SPDE implementations in \pkg{TMB} \citep{breivik2021, osgood-zimmerman2021}, penalized complexity (PC) priors \citep{simpson2017, fuglstad2019} (\texttt{?pc\_matern}) can constrain the spatial range and variance parameters.
These priors or penalties are available both with maximum likelihood estimation and with MCMC sampling.
If a model will be passed to \pkg{tmbstan} with priors, a \texttt{bayesian} logical argument should be set to \texttt{TRUE} to enable Jacobian adjustments for changes of variables (priors applied to parameters that are internally transformed) \citep{carpenter2017}.

@@

% \subsection{Software dependencies}\label{dependencies}

% \sdmTMB\ relies on several well-established \R\ packages (Figure~\ref{fig:flowchart}).
% \sdmTMB\ uses \pkg{fmesher} \citep{fmesher} and \INLA\ \citep{lindgren2015} to generate the triangulation mesh and input matrices for the SPDE calculations.
% The SPDE precision matrix calculations themselves are performed by functions within \pkg{TMB} \citep{kristensen2016} (\texttt{R\_inla} namespace).
% Data structures required to implement penalized smoothers are constructed with \texttt{smooth2random()} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
% The statistical model itself is written as a \proglang{C++} template for \TMB.
% \TMB\ returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
% The negative marginal log likelihood is minimized via the non-linear optimization routine \texttt{stats::nlminb()} \citep{gay1990} in \proglang{R} with additional optimization carried out via a Newton optimization step to find the Hessian using \texttt{stats::optimHess()} \citep{gay1990, r2021}.
% After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to samples from the joint posterior distribution for Bayesian inference.
% % TODO: visreg + ggeffects effects, etc.?

\section{Installation}\label{installation}

@@

<<sdmTMB-install, eval=FALSE, echo=TRUE>>=
install.packages("sdmTMB")
@

As an alternative to the CRAN version, the development version can be installed with:

<<remotes, eval=FALSE, echo=TRUE>>=
# install.packages("remotes")
remotes::install_github("pbs-assess/sdmTMB")
@

@@

We can test for spatial autocorrelation with a Moran's I test or with a visual inspection of the residuals calculated from the \texttt{residuals()} S3 method.
We omit example code for Moran's I calculation for brevity but include it in the code supplement.
% TODO: reword this?
We see strong evidence for spatial autocorrelation for the model without a random field (p $<$ 0.01) but no evidence for the model with a random field (p = 0.84) suggesting that residual spatial autocorrelation is alleviated by including the random field.

@@

\sloppy To visualize our model, we can make predictions with the \texttt{predict()} method (\texttt{?predict.sdmTMB}) and optionally use the \texttt{newdata} argument to predict on a new data frame containing locations and all predictor columns.
Here, we will predict on a 2 \(\times\) 2 km grid (\texttt{qcs\_grid}) that covers the entire region of interest so we can visualize the predictions spatially.
The grid contains spatial covariate columns and all predictors used in the model set at values for which we want to predict.
The output of \texttt{predict()} is a data frame containing overall estimates in link space (\texttt{est}), estimates from the non-random-field components (\texttt{est\_non\_rf}; here, intercept and depth), and estimates from the individual random field components (here, \texttt{omega\_s}---the spatial field).
We can plot these with \texttt{geom\_raster()} from the \pkg{ggplot2} package (Figure~\ref{fig:pcod-predict-maps}).

@@

% TODO component confusing:

<<dog-wcvi-pred, fig.asp = 0.8, fig.cap="Example prediction components from the spatiotemporal model of Pacific Dogfish biomass density. Throughout, two example years are shown. (a) \\code{est\\_non\\_rf1} refers to the prediction from all non-random-field components (here, a smoother for bottom depth and the time-varying year effect) from the first delta model, (b) \\code{omega\\_s1} refers to the spatial random field from the first delta model, (c) \\code{epsilon\\_st2} refers to spatiotemporal random fields from the second delta model, and (d) \\code{est} refers to the overall prediction estimate combining all effects. The spatial random field is constant through time (i.e., the two panels in b are identical) and represents static biotic or abiotic features not included as covariates (e.g., habitat). The spatiotemporal random fields are different each time step and here are allowed to follow an AR(1) process. They represent temporal variability in the spatial patterning of Pacific Spiny Dogfish (e.g., resulting from movement or local changes in population density).", fig.width=9, out.width="6in">>=
cowplot::plot_grid(
  g_nonrf,
  g_omega,
  g_eps,
  g_est,
  ncol = 2L
)
@

@@

There are many \proglang{R} packages capable of fitting geostatistical spatial or spatiotemporal models \citep[e.g.,][]{heaton2019}.
\pkg{sdmTMB}, \pkg{VAST}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, and \pkg{spaMM} \citep{rousset2014} are the most closely related, as they all provide a user interface to SPDE-based GMRF models.
In our software comparison (Table \ref{tab:functionality}), we also include \pkg{mgcv} as it can be adapted to use the SPDE \citep{miller2019} and \pkg{spBayes} \citep{finley2007, finley2015} since it is a prominent package that can fit related predictive-process models without the SPDE.
\pkg{sdmTMB}, \pkg{VAST}, and \pkg{mgcv} can estimate anisotropic covariance whereas \proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{spBayes} are currently limited to isotropic covariance.
\pkg{sdmTMB} and \pkg{mgcv} focus on univariate response data, whereas \pkg{VAST}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, \pkg{spaMM}, and \pkg{spBayes} extend to multivariate responses with various limitations.
To our knowledge, \pkg{VAST} is the only package to implement spatial \citep{thorson2015} and spatial dynamic factor analysis \citep{thorson2016} and spatial empirical orthogonal function (EOF) regression \citep{thorson2020}.
Of these packages, only \pkg{sdmTMB} and \pkg{inlabru} can currently fit threshold (e.g., hockey-stick) covariate relationships.
\pkg{spaMM} is limited to a spatial random field (i.e., does not fit spatiotemporal fields) and \pkg{spBayes} implements spatiotemporal fields, but only as a random walk.
There is considerable variability in the available observation likelihoods across packages (Table \ref{tab:functionality}).
% We provide comparisons of the syntax and the reproducibility of results from models fit using \pkg{sdmTMB}, \proglang{R}-\pkg{INLA}, and \pkg{inlabru} in Appendix~\ref{app:inla}.

@@

% TODO: figure out where to put this paragraph
% TODO: rework this paragraph to be advantages of random fields GAMs and CAR
The GLMMs underpinning \pkg{sdmTMB} models are spatially explicit---they estimate interpretable parameters of a spatial covariance function: parameters defining the magnitude of spatial variation and the rate of correlation decay with distance.
In contrast, semi- and non-parametric approaches do not estimate spatial covariance functions (e.g., \pkg{randomForest} \citep{liaw2002}, \pkg{MaxEnt} \citep{phillips2006}, and most smooths in \pkg{mgcv} \citep{wood2017a}).
% The random fields in \pkg{sdmTMB} are structured as MVN constrained by a Matérn covariance function \citep{matern1986}.
% The Matérn can accommodate a range of shapes and can be both isotropic (correlation decays the same in all directions) or anisotropic (correlation is directionally dependent) \citep{haskard2007}.
% The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to \(\sim 0.13\) \citep{lindgren2015}---can be shared among random fields or estimated separately (\texttt{share\_range} argument).
% TODO: figure out where to put this paragraph
Approaches such as conditional autoregressive models (CAR) require continuous spatial data to be aggregated to polygons or grids...
Areal data (data aggregated to polygon or grid level) may be analyzed using other spatial models, including conditional (CAR) autoregressive models \citep[e.g.,][]{verhoef2018}; \pkg{sdmTMB} can also fit models with areal data if each polygon has an associated centroid.
A benefit of the geostatistical approach over CAR or similar models is that the parameters describing spatial covariance can be more easily interpreted \citep{wall2004} (e.g., spatial variance and range---the distance at which points are effectively independent).

@@

There are several planned future additions to the \pkg{sdmTMB} model structure.
A subset of features to be added in the future include (1) multivariate models to allow for simultaneous modelling of data from multiple groups, (2) integration with the \pkg{RTMB} \citep{rtmb} package so that the model code base will be written in \R\ rather than \proglang{C++}, (3) inclusion of a dispersion formula similar to \pkg{glmmTMB},
(4) The ability to specify observation-level likelihoods to integrate different data types (e.g. REF).
(5) Conditional autoregressive (CAR) models as an alternative to random fields for large datasets that are already areal in nature.
(6) Improved visualization methods and integration

% (4) continuous-time models \citep[e.g.,][]{blangiardo2015book}, (5) non-Gaussian random fields \citep[e.g.,][]{anderson2019}, and (6) multivariate responses.
% The included \pkg{TMB} \texttt{.cpp} file is flexible and provides a tested model template that can be modified to add additional features.
% TODO random slopes looks gnarly to do generically? EW: -- agree, I mostly have looked at glmmTMB / brms implementations
% removed random slopes -- another feature that could be added is 3D models, e.g. https://doi.org/10.1016/j.spasta.2023.100750
